Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-04-18 paper.tex
Updates from Overleaf

\subsection{Variational Bound on Predictive Rate-Distortion}
We will first describe the general method, without committing to a specific framework for function approximation yet.
The idea will be to construct a bound on predictive rate-distortion and optimize this bound in a parametric family of function approximators to obtain an encoding $Z$ that is close to optimal for the nonparametric objective~(\ref{eq:ib}).

@@

Our method will be to introduce additional functions, also expressed using function approximators, that approximate some of these intractable quantities:
First, we will use a parameterized probability distribution $q$ as an approximation to the intractable marginal $P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})$:
\begin{equation}
q(Z) \text{ approximates } P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})
\end{equation}
Second, to approximate the decoding distribution $P(X_{0\dots M-1}|Z)$, we introduce a parameterized decoder $\psi$ that maps points $Z \in \mathbb{R}^N$  into probability distributions $P_\psi(X_{0\dots M-1}|Z)$ over future observations $X_{0\dots M-1}$:
\begin{equation}
P_\psi(X_{0\dots M-1}|Z)  \text{ approximates }  P(X_{0\dots M-1}|Z)
\end{equation}
for each code $Z \in \mathbb{R}^N$. Crucially, $P_\psi(X_{0\dots M-1}|Z)$ will be easy to compute efficiently, unlike the exact decoding distribution $P(X_{0\dots M-1}|Z)$.
That is, we introduce $\psi$ and $q$ as \emph{variational approximations} \citep{blei-variational-2016} to the optimal decoder and the true marginal of $Z$.

@@

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:ib-oce-repeat}) in a way amenable to using function approximators.
Note that the the predictive rate-distortion objective~(\ref{eq:ib-oce-repeat}) can equivalently be written as
\begin{equation}\label{eq:prd-recall}
	\min_{Z : Z \bot X_{0\dots M-1} | X_{-M\dots-1}}	\left[\operatorname{H}[X_{0\dots M-1} | Z] + \lambda \operatorname{I}[ X_{-M\dots-1}, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
By the two Propositions, this objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] + \lambda \cdot \operatorname{D_{KL}}\infdivx{P_\phi(\cdot|X_{-M\dots-1})}{q}\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-recall}) by setting $Z$ to be the code generated by the encoder $\phi$.
The second proposition guarantees that, for a solution to  (\ref{eq:prd-recall}), the rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] =   \mathbb{E}_{X_{-M\dots-1}}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots-1})}{q(Z)}\right]
\end{equation}
and the predictiveness is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{I}[Z, X_{0\dots M-1}] =  \operatorname{H}[X_{0\dots M-1}] - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] %H[X_{0\dots M-1}|X_{-M\dots-1}]
\end{equation}
Note that the right-hand-sides of (\ref{eq:eq1-new}-\ref{eq:eq2-new}) can both be estimated efficiently using Monte Carlo samples from $Z \sim P_\phi(X_{-M\dots-1})$.

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and predictiveness for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[X_{0\dots M-1}|X_{-M\dots-1}]
  \operatorname{I}[Z, X_{0\dots M-1}] \geq  \operatorname{H}[X_{0\dots M-1}] - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and -- if the family of approximators is sufficiently rich---can come close to turning these into the equalities  (\ref{eq:eq1-new}-\ref{eq:eq2-new}).

\subsection{Implementation}
\paragraph{Approximating Families}
We choose the approximating families for the encoder $\phi$, the decoder $\psi$, and the distribution $q$ to be certain types of neural networks.

For $\phi$ and $\psi$, we use recurrent neural networks with LSTM cells \citep{hochreiter-long-1997}, widely used for modeling sequential data across different domains.
We parameterize the distribution $P_\phi(Z|X_{-M\dots-1})$ as a Gaussian whose mean and variance are computed from the past $X_{-M\dots-1}$:
We use an LSTM network to compute a vector $h \in \mathbb{R}^k$ from the past observations $X_{-M\dots-1}$, and then compute
\begin{equation}\label{eq:z-gauss}
	Z \sim \mathcal{N}(W_\mu h, (W_\sigma h)^2 I_{k\times k})
\end{equation}
where $W_\mu, W_\sigma \in \mathbb{R}^{k\times k}$ are parameters.
While we found Gaussians sufficiently flexible for $\phi$, more powerful encoders could be constructed using more flexible parametric families, such as normalizing flows \citep{rezende-variational-2015, kingma-improving-2016}.

For the decoder $\psi$, we use a second LSTM network to compute a sequence of vector representations $g_t = \psi(Z, X_{1\dots t-1})$ ($g_t \in \mathbb{R}^k$) for $t = 1, \dots T$.
We compute predictions using the softmax rule 
\begin{equation}
	P_\psi(X_t = s_i|X_{1...t-1}, Z) \propto \exp((W_o g_t)_i)
\end{equation}
for each element $s_i$ of the state space $S = \{s_1, ..., s_{|S|}\}$, and $W_o \in \mathbb{R}^{|S| \times k}$ is a parameter matrix to be optimized together with the parameters of the LSTM networks.

@@

\paragraph{Optimization and Evaluation}
We optimize the parameters of the neural networks expressing $\phi, \psi, q$ for the objective (\ref{eq:bound}) using Backpropagation and Adam \citep{kingma-adam:-2014}, a standard and widely used gradient descent-based method for optimizing neural networks.
During optimization, we approximate the gradient by taking a single sample from $Z$ (\ref{eq:z-gauss}) per sample trajectory $X_{-M}, \dots, X_{M-1}$ and use the reparametrized gradient estimator introduced by \citet{kingma-auto-encoding-2014}.
This results in an unbiased estimator of the gradient of (\ref{eq:bound}) w.r.t. the parameters of $\phi, \psi, q$.

@@

%\paragraph{Evaluation}
After optimizing the parameters on a set of observed trajectories, we estimate rate and prediction error on the test set.
Given parameters for $\phi, \psi, q$, we evaluate the PRD objective~(\ref{eq:bound}), rate (\ref{eq:bound1-new}), and the prediction error (\ref{eq:bound2-new}) on the test set by taking, for each time series $X_{-M}...X_{M-1}$ from the test set, a single sample $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing Monte Carlo estimates for rate
\begin{equation}\label{eq:bound-mc-rate}
\mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right] \approx	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}  \log \frac{p_{\mathcal{N}(\mu, \sigma^2)}(z)}{q(z)}
\end{equation}
where $p_{\mathcal{N}(\mu, \sigma^2)}(z)$ is the Gaussian density with $\mu, \sigma^2$ computed from $X_{-M\dots -1}$ as in (\ref{eq:z-gauss}), and prediction error 
\begin{equation}\label{eq:bound-mc-distortion}
 -	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] \approx \frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P_\psi(X_{0\dots M-1} | Z)
\end{equation}
%and the PRD objective
%\begin{equation}\label{eq:bound-mc}
%	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
%\end{equation}
%Note that one can go from distortion to predictiveness by estimating
Thanks to (\ref{eq:bound1-new}-\ref{eq:bound2-new}), these estimates are guaranteed to be \emph{upper bounds} on the true rate and prediction error achieved by the code $Z  \sim \mathcal{N}(\mu, \sigma^2)$, up to sampling error introduced into the Monte Carlo estimation by sampling $z$ and the finite size of the test set.

@@

\subsection{Related Work}
In (\ref{eq:bound}), we derived a variational formulation of predictive rate-distortion.
This is formally related to a variational formulation of the Information Bottleneck that was introduced by \cite{alemi-deep-2016}, who applied it to neural-network based image recognition.
Unlike our approach, they used a fixed diagonal Gaussian instead of a parametrized distribution for $q$.
As far as we are aware, no previous work has used such a variational formulation in the modeling of sequences.

@@

@@

\section{Experiments}\label{sec:experiments}
We now test the ability of NPRD to estimate rate-distortion curves.
Before we apply NPRD to obtain the first estimates of predictive rate-distortion for natural language in Section~\ref{sec:language}, we validate the method on processes whose tradeoff curves are analytically known, and compare with OCE.

@@

\paragraph{OCE}
As discussed in Section~\ref{sec:oce}, OCE overestimates structure in the process due to overfitting, %resulting in biased estimates of rate and distortion.More specifically, 
overfitting will result in an overly optimistic estimate, systematically overestimating the predictiveness achieved at a given rate.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and distortion on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|X_{-M\dots -1})$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction loss.
This method not only enables fair comparison between OCE and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{ineq2}:
\begin{equation}\label{eq:rate-oce}
\frac{1}{N} \sum_{X_{-M}\dots X_{-1} \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|X_{-M}\dots X_{-1})}{s(Z)}
\end{equation}
where $P(Z|X_{-M}\dots X_{-1})$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-oce}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{ineq1}:
\begin{equation}
\frac{1}{N} \sum_{X_{-M}\dots X_{M-1} \in \text{Test Data}} \E_{Z \sim P(\cdot|X_{-M \dots -1})} \log P(X_{0\dots M-1}|Z).
\end{equation}
where $P(X_{0\dots M-1}|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to (\ref{ineq1}-\ref{ineq2}), these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

\paragraph{Recovering Causal States}
Does NPRD lead to interpretable codes $Z$?
To answer this, we further investigated the NPRD approximation to the Random Insertion Process (RIP), obtained in the previous paragraph.
The $\epsilon$-machine was computed by  \cite{marzen-predictive-2016} and is given in Figure~\ref{fig:rip-machine} (left).
The process has three causal states:
A represents those pasts where the future starts with $1^{2k}0$ ($k = 0, 1, 2, \dots$) -- these are the pasts ending in either $001$ or $10111^m$ ($m= 0, 1, 2, \dots$). % but cannot go in with 10
B represents those pasts ending in 10 -- the future has to start with 01 or 11.
C represents those pasts ending in either 00 or 01 -- the future has to start with $1^{2k+1}0$ ($k = 0, 1, 2, \dots$).

@@

\subsection{Discussion}
Let us now interpret the curves in Figure~\ref{fig:eng-pos}. \rljf{What is predictiveness? Excess entropy?}
The rate-predictiveness curve (left) shows that, at low rates, predictiveness is approximately proportional to the rate.
At greater degrees of predictiveness, rate grows faster and faster, whereas predictiveness seems to asymptote to $\approx 1.1$ nats.
%(In the case of our experiment, since $T = 15$, rate would be bounded by the aggregate block-entropy of 15 words, about 28 nats.)
The asymptote of predictiveness can be identified with the excess entropy $E_0 := \operatorname{I}[X_{-M\dots -1}, X_{0 \dots M-1}]$, whereas the rate should asymptote to the statistical complexity.
Judging by the curve, natural language---at the level we are measuring in this experiment---has a statistical complexity much higher than its excess entropy:
At the highest rate measured by NPRD in our experiment, rate is about 20 nats, whereas predictiveness is about 1.1 nats.
If these values are correct, then---due to the convexity of the rate-predictivity curve---statistical complexity exceeds the excess entropy by a factor of at least $\frac{20}{1.1}$. \rljf{Wow, is this true for any other natural processes?}
Note that this picture agrees qualitatively with the OCE results, which suggest a lower-bound on the ratio of at least $\frac{2.5}{0.6} > 5$.

@@

Note that there are many possible ways of fitting these curves; we have described a simple one that requires two free parameters $\alpha >0$, $\beta > 1$, in addition to a guess for the maximal predictiveness $E_0$.
In any case, the results show that natural language shows an approximately linear growth of predictiveness with rate at small rates, and exploding rates at diminishing returns in predictiveness later.
Why might language show such behavior?
%We will describe a proposal:
One natural idea is that the causal states of natural language are the singletons $\overleftarrow{X}$, that is, no two past trajectories induce a fully equivalent distribution over future trajectories.
Then the code $Z^{(T)}$ which stores the block $X_{-T...-1}$ has rate $R_T = \Theta(T)$, but its predictiveness $P_T =	\operatorname{I}[X_{-T...-1}, X_{0,...M-1}]$ asymptotes to $E_0$.

@@

\section{Conclusion}
We introduce Neural Predictive Rate Distortion (NPRD), a method for estimating predictive rate-distortion when only sample trajectories are given.
Unlike OCE, the most general prior method, NPRD scales to long sequences and large state spaces.
On analytically tractable processes, we show that it closely fits the analytical rate-distortion curve and recovers the causal states of the process.
On POS-level modeling of natural language, it agrees with OCE in the setting of low rate and short sequences; outside these settings, OCE fails due to combinatorial explosion and overfitting, while NPRD continues to provide estimates.
Finally, we use NPRD to provide the first estimates of predictive rate-distortion for modeling natural language.
%Results suggest that prediction can be approximated with sim, but more accurate prediction requires very fine-grained memory representations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\funding{This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''. Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}

>>>>

\subsection{Variational Bound on Predictive Rate-Distortion}
We will first describe the general method, without committing to a specific framework for function approximation yet.
We will construct a bound on predictive rate-distortion and optimize this bound in a parametric family of function approximators to obtain an encoding $Z$ that is close to optimal for the nonparametric objective~(\ref{eq:ib}).

@@

% That is, we introduce $\psi$ and $q$ as \emph{variational approximations} \citep{blei-variational-2016} to the optimal decoder and the true marginal of $Z$.

Our method will be to introduce additional functions, also expressed using function approximators, that approximate some of these intractable quantities:
First, we will use a parameterized probability distribution $q$ as an approximation to the intractable marginal $P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})$:
\begin{equation}
q(Z) \text{ approximates } P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})
\end{equation}
Second, to approximate the decoding distribution $P(X_{0\dots M-1}|Z)$, we introduce a parameterized decoder $\psi$ that maps points $Z \in \mathbb{R}^N$  into probability distributions $P_\psi(X_{0\dots M-1}|Z)$ over future observations $X_{0\dots M-1}$:
\begin{equation}
P_\psi(X_{0\dots M-1}|Z)  \text{ approximates }  P(X_{0\dots M-1}|Z)
\end{equation}
for each code $Z \in \mathbb{R}^N$. Crucially, $P_\psi(X_{0\dots M-1}|Z)$ will be easy to compute efficiently, unlike the exact decoding distribution $P(X_{0\dots M-1}|Z)$.

@@

\begin{proof}
The two equalities follow from the definition of KL Divergence and Mutual Information. To show the inequality, we again use Gibbs' inequality:
\begin{align*}
    -\mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log q(Z) = - \mathbb{E}_Z \log q(Z) \geq - \mathbb{E}_Z \log P(Z) = -\mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log P(Z) 
\end{align*}
Here, equality holds if and only if $q(Z) = P(Z)$, proving the second assertion.
\end{proof}

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:ib-oce-repeat}) in a way amenable to using function approximators.
Note that the predictive rate-distortion objective~(\ref{eq:ib-oce-repeat}) can equivalently be written as
\begin{equation}\label{eq:prd-recall}
	\min_{Z : Z \bot X_{0\dots M-1} | X_{-M\dots-1}}	\left[\operatorname{H}[X_{0\dots M-1} | Z] + \lambda \operatorname{I}[ X_{-M\dots-1}, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
By the two Propositions, this objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] + \lambda \cdot \operatorname{D_{KL}}\infdivx{P_\phi(\cdot|X_{-M\dots-1})}{q}\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-recall}) by setting $Z$ to be the code generated by the encoder $\phi$.
The second proposition guarantees that, for a solution to  (\ref{eq:prd-recall}), the rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] =   \mathbb{E}_{X_{-M\dots-1}}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots-1})}{q(Z)}\right]
\end{equation}
and the prediction error is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[X_{0\dots M-1}|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] %H[X_{0\dots M-1}|X_{-M\dots-1}] % \operatorname{H}[X_{0\dots M-1}]
\end{equation}
Note that the right-hand-sides of (\ref{eq:eq1-new}-\ref{eq:eq2-new}) can both be estimated efficiently using Monte Carlo samples from $Z \sim P_\phi(X_{-M\dots-1})$.

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and prediction error for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[X_{0\dots M-1}|X_{-M\dots-1}]
  \operatorname{H}[X_{0\dots M-1}|Z] \leq   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] % \operatorname{H}[X_{0\dots M-1}]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and -- if the family of approximators is sufficiently rich---can come close to turning these into the equalities  (\ref{eq:eq1-new}-\ref{eq:eq2-new}).

\subsection{Choosing Approximating Families}
For our method of Neural Predictive Rate Distortion (NPRD), we choose the approximating families for the encoder $\phi$, the decoder $\psi$, and the distribution $q$ to be certain types of neural networks that are known to provide strong and general models of sequences and distributions.

For $\phi$ and $\psi$, we use recurrent neural networks with Long Short Term Memory (LSTM) cells \citep{hochreiter-long-1997}, widely used for modeling sequential data across different domains.
We parameterize the distribution $P_\phi(Z|X_{-M\dots-1})$ as a Gaussian whose mean and variance are computed from the past $X_{-M\dots-1}$:
We use an LSTM network to compute a vector $h \in \mathbb{R}^k$ from the past observations $X_{-M\dots-1}$, and then compute
\begin{equation}\label{eq:z-gauss}
	Z \sim \mathcal{N}(W_\mu h, (W_\sigma h)^2 I_{k\times k})
\end{equation}
where $W_\mu, W_\sigma \in \mathbb{R}^{k\times k}$ are parameters.
While we found Gaussians sufficiently flexible for $\phi$, more powerful encoders could be constructed using more flexible parametric families, such as normalizing flows \citep{rezende-variational-2015, kingma-improving-2016}.

For the decoder $\psi$, we use a second LSTM network to compute a sequence of vector representations $g_t = \psi(Z, X_{0\dots t-1})$ ($g_t \in \mathbb{R}^k$) for $t = 0, \dots M-1$.
We compute predictions using the softmax rule 
\begin{equation}
	P_\psi(X_t = s_i|X_{1...t-1}, Z) \propto \exp((W_o g_t)_i)
\end{equation}
for each element $s_i$ of the state space $S = \{s_1, ..., s_{|S|}\}$, and $W_o \in \mathbb{R}^{|S| \times k}$ is a parameter matrix to be optimized together with the parameters of the LSTM networks.

@@

\subsection{Parameter Estimation and Evaluation}
We optimize the parameters of the neural networks expressing $\phi, \psi, q$ for the objective (\ref{eq:bound}) using Backpropagation and Adam \citep{kingma-adam:-2014}, a standard and widely used gradient descent-based method for optimizing neural networks.
During optimization, we approximate the gradient by taking a single sample from $Z$ (\ref{eq:z-gauss}) per sample trajectory $X_{-M}, \dots, X_{M-1}$ and use the reparametrized gradient estimator introduced by \citet{kingma-auto-encoding-2014}.
This results in an unbiased estimator of the gradient of (\ref{eq:bound}) w.r.t. the parameters of $\phi, \psi, q$.

@@

%\paragraph{Evaluation}
After optimizing the parameters on a set of observed trajectories, we estimate rate and prediction error on the test set.
Given parameters for $\phi, \psi, q$, we evaluate the PRD objective~(\ref{eq:bound}), rate (\ref{eq:bound1-new}), and the prediction error (\ref{eq:bound2-new}) on the test set by taking, for each time series $X_{-M}...X_{M-1}$ from the test set, a single sample $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing Monte Carlo estimates for rate
\begin{equation}\label{eq:bound-mc-rate}
\mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right] \approx	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}  \log \frac{p_{\mathcal{N}(\mu, \sigma^2)}(z)}{q(z)}
\end{equation}
where $p_{\mathcal{N}(\mu, \sigma^2)}(z)$ is the Gaussian density with $\mu, \sigma^2$ computed from $X_{-M\dots -1}$ as in (\ref{eq:z-gauss}), and prediction error 
\begin{equation}\label{eq:bound-mc-distortion}
 -	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] \approx - \frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P_\psi(X_{0\dots M-1} | Z)
\end{equation}
%and the PRD objective
%\begin{equation}\label{eq:bound-mc}
%	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
%\end{equation}
%Note that one can go from distortion to predictiveness by estimating
Thanks to (\ref{eq:bound1-new}-\ref{eq:bound2-new}), these estimates are guaranteed to be \emph{upper bounds} on the true rate and prediction error achieved by the code $Z  \sim \mathcal{N}(\mu, \sigma^2)$, up to sampling error introduced into the Monte Carlo estimation by sampling $z$ and the finite size of the test set.

@@

\subsection{Related Work}
In (\ref{eq:bound}), we derived a variational formulation of predictive rate-distortion.
This is formally related to a variational formulation of the Information Bottleneck that was introduced by \cite{alemi-deep-2016}, who applied it to neural-network based image recognition.
Unlike our approach, they used a fixed diagonal Gaussian instead of a flexible parametrized distribution for $q$.
As far as we are aware, no previous work has used such a variational formulation in the modeling of sequences.

@@

%More broadly, our method is related to  The idea will be to use Variational Inference~\citep{blei-variational-2016}:

@@

\section{Experiments}\label{sec:experiments}
%In Section~\ref{sec:nprd}, we described our method Neural Predictive Rate Distortion (NPRD) 
We now test the ability of our new method NPRD to estimate rate-distortion curves.
Before we apply NPRD to obtain the first estimates of predictive rate-distortion for natural language in Section~\ref{sec:language}, we validate the method on processes whose tradeoff curves are analytically known, and compare with OCE.

@@

\paragraph{OCE}
As discussed in Section~\ref{sec:oce}, OCE overestimates structure in the process due to overfitting, %resulting in biased estimates of rate and distortion.More specifically, 
overfitting will result in an overly optimistic estimate, systematically overestimating the predictiveness achieved at a given rate.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and distortion on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|X_{-M\dots -1})$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction error.
This method not only enables fair comparison between OCE and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{ineq2}:
\begin{equation}\label{eq:rate-oce}
\frac{1}{N} \sum_{X_{-M}\dots X_{-1} \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|X_{-M}\dots X_{-1})}{s(Z)}
\end{equation}
where $P(Z|X_{-M}\dots X_{-1})$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-oce}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction error on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{ineq1}:
\begin{equation}
\frac{1}{N} \sum_{X_{-M}\dots X_{M-1} \in \text{Test Data}} \E_{Z \sim P(\cdot|X_{-M \dots -1})} \log P(X_{0\dots M-1}|Z).
\end{equation}
where $P(X_{0\dots M-1}|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to (\ref{ineq1}-\ref{ineq2}), these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

\paragraph{Recovering Causal States}
Does NPRD lead to interpretable codes $Z$?
To answer this, we further investigated the NPRD approximation to the Random Insertion Process (RIP), obtained in the previous paragraph.
The $\epsilon$-machine was computed by  \cite{marzen-predictive-2016} and is given in Figure~\ref{fig:rip-machine} (left).
The process has three causal states:
State \textit{A} represents those pasts where the future starts with $1^{2k}0$ ($k = 0, 1, 2, \dots$) -- these are the pasts ending in either $001$ or $10111^m$ ($m= 0, 1, 2, \dots$). % but cannot go in with 10
State \textit{B} represents those pasts ending in 10 -- the future has to start with 01 or 11.
State \textit{C} represents those pasts ending in either 00 or 01 -- the future has to start with $1^{2k+1}0$ ($k = 0, 1, 2, \dots$).

@@

\subsection{Discussion}
Let us now consider the curves in Figure~\ref{fig:eng-pos} in more detail.
The rate-predictiveness curve (left) shows that, at low rates, predictiveness is approximately proportional to the rate.
At greater degrees of predictiveness, rate grows faster and faster, whereas predictiveness seems to asymptote to $\approx 1.1$ nats.
The asymptote of predictiveness can be identified with the mutual information between past and future observations, $E_0 := \operatorname{I}[X_{-M\dots -1}, X_{0 \dots M-1}]$, which is a lower bound on the excess entropy. The rate should asymptote to the statistical complexity.
Judging by the curve, natural language---at the time scale we are measuring in this experiment---has a statistical complexity much higher than its excess entropy:
At the highest rate measured by NPRD in our experiment, rate is about 20 nats, whereas predictiveness is about 1.1 nats.
If these values are correct, then---due to the convexity of the rate-predictivity curve---statistical complexity exceeds the excess entropy by a factor of at least $\frac{20}{1.1}$. \rljf{Wow, is this true for any other natural processes?}
Note that this picture agrees qualitatively with the OCE results, which suggest a lower-bound on the ratio of at least $\frac{2.5}{0.6} > 5$.

@@

Note that there are other possible ways of fitting these curves; we have described a simple one that requires only two parameters $\alpha >0$, $\beta > 1$, in addition to a guess for the maximal predictiveness $E_0$.
In any case, the results show that natural language shows an approximately linear growth of predictiveness with rate at small rates, and exploding rates at diminishing returns in predictiveness later.
Why might language show such behavior?
%We will describe a proposal:
One natural idea is that the causal states of natural language are the singletons $\overleftarrow{X}$, that is, no two past trajectories induce a fully equivalent distribution over future trajectories.
Then the code $Z^{(T)}$ which stores the block $X_{-T...-1}$ has rate $R_T = \Theta(T)$, but its predictiveness $P_T =	\operatorname{I}[X_{-T...-1}, X_{0,...M-1}]$ asymptotes to $E_0$.

@@

\section{Conclusion}
We introduce Neural Predictive Rate Distortion (NPRD), a method for estimating predictive rate-distortion when only sample trajectories are given.
Unlike OCE, the most general prior method, NPRD scales to long sequences and large state spaces.
On analytically tractable processes, we show that it closely fits the analytical rate-distortion curve and recovers the causal states of the process.
On part-of-speech-level modeling of natural language, it agrees with OCE in the setting of low rate and short sequences; outside these settings, OCE fails due to combinatorial explosion and overfitting, while NPRD continues to provide estimates.
Finally, we use NPRD to provide the first estimates of predictive rate-distortion for modeling natural language in five different languages, finding qualitatively very similar curves in all languages.
%Results suggest that prediction can be approximated with sim, but more accurate prediction requires very fine-grained memory representations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\funding{This research received no external funding.} %'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''. Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}
